Allowing Stable Groups of Prosumers to Enter the Market through Proper Aggregation

### 2014-08-26 Article.tex
update

\section{Introduction}
\label{sec:introduction}
% no \IEEEPARstart
One of the most basic ideas in the smart grid revolves around the introduction of communication means in the power grids, which could enable complex improvements in the energy management and lead progressively to a greener energetic system [1] [2]. Distributed energy resources (DER) such as wind turbines or photovoltaic panels are supposed to emerge and populate not only spread out farms, but also classical residential neighborhoods. Together with electric vehicles, and demand side management tools, they constitute the building blocks which will help turn the today pure energy consumers into true actors of the grid operation [3]. Such energy aware agents that consume, produce, and are ready to make concessions (appliances delays, V2G…) to ensure grid stability are commonly called prosumers \cite{Rathnayaka2012}.

It is rather clear that there should be communication flows between prosumers and the grid as most of the demand side management concepts suppose such a link [6] [7]. Nevertheless as the expected number of active prosumers in the future seems quite large, the literature assume more and more communication between prosumers as this is a key ingredient to maintain more complex structures (such as coalitions), and thereby decrease the communication load on the grid’s side [8]. Furthermore, it has been shown that coalitions in the smart grid provide nice ways of improving stability while posing exciting challenges in terms of network infrastructure and management. Self sufficient microgrids that can disconnect from the main grid [9], virtual power plants (VPP) that map small DER in sizeable and adjustable plants [10], or coalitions of electric vehicles that back up the grid in emergency situations (V2G) [11] are just a few examples of how coalitional concepts can enhance grid reliability.

In this paper, our focus is on how coalitions of prosumers, aiming at selling energy to the grid, should be formed in order to support it efficiently. Namely, improving statistically production stability while decreasing communication needs for staying in stable state. The idea is that, by using prediction techniques, coalitions can propose contract values to the grid, enabling it to schedule production on its side accordingly. Actually, contract values are particularly relevant in power exchanges contexts where energy is traded based on predictions (day-ahead estimations for instance). It seems obvious that, in such systems, participants should try to minimize prediction errors in order to maintain a stable state of grid operation, and that some penalty rules should be applied to ensure that coalitions are willing to report contract values correctly. Furthermore, as the coalition productions are supposed to come from (uncertain) renewables, the notion of reliability comes into play with a visible tradeoff between contract values worth and reliability.

@@

In more details, and as will be explained in section \ref{sec:model}, we consider agents that, depending on meteorological conditions, personal preferences, and their appliances (loads and generators), consume and produce more or less energy (using only wind turbines and PV as generators, which can be quite simply and efficiently modeled).  We used real meteorological traces [13] to account for seasonal and daily variations in the prosumers energetic output profiles as well as realistic correlations. With these ingredients, we are able to run simulations and record the different output profiles as timeseries.

As correlations play a significant role in our mechanism, we organize the prosumers in a correlation graph, a quite popular approach in market stocks clustering [16] [17] (see section \ref{sec:related}). We will see that in such graphs, cliques represent structures of interest. In our settings, it means that cliques are “good” (from the utility point of view) foundations for high valued coalitions. We thereby use a modified clique percolation algorithm (see section \ref{sec:results}) that will enable us to expand the cliques as much as needed in order to form proper coalitions that fulfill grid requirements.

@@

In \cite{Kota2011}, the authors study the formation of virtual power plants (VPP) composed of multiple self-interested DER. On the grid side, two requirements for the formation of virtual power plants are derived, namely the reliability of supply and the minimization of entities the grid has to deal with. From this, \cite{Kota2011} builds a pricing mechanism that encourages VPP to report true estimates of their aggregated production and penalizes prediction errors. A redistribution scheme of the VPP to the DER is also constructed such that the payoff allocation lies in the core of the game, meaning that no DER has an incentive to leave the coalition.

In this paper and as will be explained in section \ref{sec:model}, coalitions utilities depend on statistical properties of historical values. The goal being to form coalitions that, statistically spealing, are more likely to exhibit stable behaviors. We will see that, when it comes to the formation of the coalitions, a prosumer is completely interchangeable with the timeserie representing its available production.

@@

The following section of this paper explains the model we built in order obtain sufficient historical values for our prosumers, such that forming coalitions based on statistics is actually meaningfull.

@@

We now choose a very simple utility function that derives directly from the above remarks. If a coalition cannot provide a valid contract value, it receives naturally a utility of zero. Furthermore, it seems obvious that the utility should increase with the contract value. The $ 1/|S| $ term indicates that we favorise small coalitions, mainly because they are easier to maintain in term of communication.

@@

The strategy of $ \{ij\} $ consists in maximizing its contract values as long as it respects this inequality, e.g to annonce $ P_{\phi}(ij) = \mu_{ij} - \sigma_{ij}\sqrt{2}erf^{-1}(1-2 \phi ) $. It thus appears (as it was intuitively understandable) that, for equivalent sizes, coalitions with low relative standard deviations ( $ \sigma_{ij} / \mu_{ij} $ ) are able to announce higher contract values.

@@

The model we used in order to simulate timeseries of available production provide some diversity because of the energetic mix/climate vectors combo. Nevertheless, as the number of agents grows, the timeseries tend to exhibit similar patterns. This is particularly visible when creating a correlation graph $ G_{1}(\mathcal{A},E_{1},\omega_{1}) $ with the same kind of metric as \cite{Garas2008} or \cite{Onnela2004} ($ d_{ij} = 1 - \rho_{ij}^{2} $). Very well defined clusters appear in the corresponding $ \epsilon $-graph for large range of $ \epsilon $.

@@

We thus opt for reversing the metric ($ d_{ij}^{'} = \rho_{ij}^{2} $) such that uncorrelated timeseries are close to each other and (anti)correlated timeseries are distant in $ G_{2}(\mathcal{A},E_{2},\omega_{2}) $. As expected \cite{Onnela2004}, independently of the $ \epsilon $ parameter, the resulting $ \epsilon $-graph exhibits henceforth much less clustering than in $ G_{1} $ and communities seem hardly visible. Therefore, using classical clustering or community detection algorithms do not provide very good results.

@@

In the following, we fix the reliability to a given empirical value and observe how the coalitions evolves for different values of $P_{\phi}^{MIN}$. Note that the opposite is also possible although a little less intuitive, but not shown here because it leads to similar conclusions.

On the right plot of figure \ref{Fig2}, we can see that when the grid constraints are fixed ($ \phi = 0.08 $ and $ P_{\phi}^{MIN} = 2 $ for instance) and only the number of coalitions varies, there is an increase in social welfare (sum of coalitions utilities) up to a maximum point (for $ N_{COAL} = 3 $ here) before a decrease. The reason is that increasing the number of coalitions allows more coalitions to reach stability and enter the market, but there is a point where nodes bringing stability are not sufficient inside the coalitions to make them pass the grid requirements, and some coalitions start to fail with zero utility. Moreover, reckon that increasing $ N_{COAL} $ means also increasing $ \epsilon $, leading to denser graphs where information is flooded, meaning that the algorithm performances decrease.
Naturally, it is also visible that small values of $P_{\phi}^{MIN}$ lead to higher utilities because coalitions are able to announce higher contract values.

\begin{center}
\begin{figure}[b]
  \includegraphics[scale=0.8]{./figure5/coals.pdf}
  \caption{Evolution of the coalitions for different values of grid requirement $ P_{MIN} $. Blue dots represents clique percolation results while red dots stands for random process results.}
  \label{Fig3}
\end{figure}
\end{center}

For our 200 prosumers, we suppose that the grid fixes a reliability $ \phi = 0.1 $ (meaning that coalitions should produce more than their contract values at least $ 90 \% $ of the time). We also fix the number of coalitions to 10 and have a look at how these coalitions evolve as the grid changes the $ P_{MIN} $ requirement. As a comparison, we use a completely random algorithm that only asks for a number of coalitions and partitions the agents in a random fashion. Figure \ref{Fig3} shows this evolution for our algorithm (blue dots) and for the random process (red dots). The diameter of a dot is proportional to the number of agents in the coalition and the higher the dot is, the higher its utility. The $ P_{MIN} $ values of the x axis are expressed in tenth of MW for readability and the hatched zone corresponds to the "under-requirement space", meaning that whenever a coalition is in this zone, it has a null utility.

Looking at figure \ref{Fig3}, we can see first that our algorithm seems to make generally a better job at finding high valued coalitions, but also that the results seem far more robust against how the grid positions its requirements. The blue dots allowed to enter the market seem indeed to outnumber the red dots, especially when the grid requirements are neither to low or to high ($ P_{MIN} = 5 $ in figure \ref{Fig3}).

In order to explore a little bit more in this direction, we have a look, on figure \ref{Fig4}, at the percentage of coalitions able to enter the market for different values of $ P_{MIN} $. For consistency, we always average the results of the random procedure over 100 realizations (5 realizations for our algorithm, but the results are relatively stable), and the errorbars in the plots stands for the standard deviations of the results.

When the grid requirements are very low ( $ P_{MIN} $ close to zero), all coalitions for both algorithms are able to enter the market, yielding an acceptance percentage of $ 100 \% $. But as $ P_{MIN} $ increases, we see the percentage of the random procedure quickly dropping while it stays constant for our algorithm. For $ P_{MIN} = 4 $, we see that only a little more than half of the coalitions for the random case are able to enter while all of them enters for the clique percolation algorithm. Finally, when the grid requirement becomes to high, the acceptance percentage of our algorithm tends slowly to zero (only partly shown in this plot for readability).

\begin{center}
\begin{figure}
   \includegraphics[scale=0.7]{./figure4/survival.eps}
   \caption{Percentage of coalitions able to enter the market for different values of grid requirement $ P_{MIN} $. Blue curve represents clique percolation results while red curve stands for random process results}
   \label{Fig4}
\end{figure}
\end{center}

>>>>

\section{Introduction}
\label{sec:introduction}
% no \IEEEPARstart
One of the most basic ideas in the smart grid revolves around the introduction of communication means in the power grids, which could enable complex improvements in the energy management and lead progressively to a greener energetic system \cite{Ramchurn} \cite{WuHamedHuangBook2011}. Distributed energy resources (DER) such as wind turbines or photovoltaic panels are supposed to emerge and populate not only spread out farms, but also classical residential neighborhoods. Together with electric vehicles, and demand side management tools, they constitute the building blocks which will help turn the today pure energy consumers into true actors of the grid operation \cite{Ramchurn}. Such energy aware agents that consume, produce, and are ready to make concessions (appliances delays, V2G…) to ensure grid stability are commonly called prosumers \cite{Rathnayaka2012} \cite{Ramchurn}.

It is rather clear that there should be communication flows between prosumers and the grid as most of the demand side management concepts suppose such a link \cite{WuHamedHuangBook2011}. Nevertheless as the expected number of active prosumers in the future seems quite large, the literature assume more and more communication between prosumers as this is a key ingredient to maintain more complex structures (such as coalitions), and thereby decrease the communication load on the grid’s side. Furthermore, it has been shown that coalitions in the smart grid provide nice ways of improving stability while posing exciting challenges in terms of network infrastructure and management. Self sufficient microgrids that can disconnect from the main grid \cite{Pahwa}, virtual power plants (VPP) that map small DER in sizeable and adjustable plants \cite{Braun} \cite{Ramchurn}, or coalitions of electric vehicles that back up the grid in emergency situations (V2G) \cite{Ramchurn} are just a few examples of how coalitional concepts can enhance grid reliability.

In this paper, our focus is on how coalitions of prosumers, aiming at selling energy to the grid, should be formed in order to support it efficiently. Namely, improving statistically the production stability while decreasing the communication needs for staying in a stable state. The idea is that, by using prediction techniques, coalitions can propose contract values to the grid, enabling it to schedule production on its side accordingly. Actually, contract values are particularly relevant in power exchanges contexts where energy is traded based on predictions (day-ahead estimations for instance). It seems obvious that, in such systems, participants should try to minimize prediction errors in order to maintain a stable state of grid operation, and that some penalty rules should be applied to ensure that coalitions are willing to report contract values correctly. Furthermore, as the coalition productions are supposed to come from (uncertain) renewables, the notion of reliability comes into play with a visible tradeoff between contract values worth and reliability.

@@

In more details, and as will be explained in section \ref{sec:model}, we consider agents that, depending on meteorological conditions, personal preferences, and their appliances (loads and generators), consume and produce more or less energy (using only wind turbines and PV as generators, which can be quite simply and efficiently modeled).  We used real meteorological traces (see section \ref{sec:model}) to account for seasonal and daily variations in the prosumers energetic output profiles as well as realistic correlations. With these ingredients, we are able to run simulations and record the different output profiles as timeseries.

As correlations play a significant role in our mechanism, we organize the prosumers in a correlation graph, a quite popular approach in market stocks clustering since \cite{Mantegna1999} (see section \ref{sec:related}). We will see that in such graphs, cliques represent structures of interest. In our settings, it means that cliques are “good” (from the utility point of view) foundations for high valued coalitions. We thereby use a modified clique percolation algorithm (see section \ref{sec:results}) that will enable us to expand the cliques as much as needed in order to form proper coalitions that fulfill grid requirements.

@@

Attacking the stability issues of renewable DER, the TradeWind project \cite{Europe} simulated the impact of wind power on electricity exchange and cross-border congestions by using a flow-based market model. The idea revolves around identifying key european interconnections (already existant or not) in order "to make optimal use of the european spacial de-correlation of wind power". It was indeed shown that geographical aggregation provides smoothening effects and that the amount of prediction errors for wind power in a geographical region diminishes as the region size increases, especially for short forecast horizons.

On a narrower scale, the authors of \cite{Kota2011} study the formation of virtual power plants (VPP) composed of multiple self-interested DER. On the grid side, two requirements for the formation of virtual power plants are considered, namely the reliability of supply and the minimization of entities the grid has to deal with. From this, \cite{Kota2011} builds a pricing mechanism that encourages VPP to report true estimates of their aggregated production and penalizes prediction errors. A redistribution scheme of the VPP to the DER is also constructed such that the payoff allocation lies in the core of the game, meaning that no DER has an incentive to leave the coalition.

In this paper and as will be explained in section \ref{sec:model}, coalitions utilities depend on statistical properties of historical values. The goal being to form coalitions that, statistically speaking, are more likely to exhibit stable behaviors. We will see that, when it comes to the formation of the coalitions, a prosumer is completely interchangeable with the timeserie representing its available production.

@@

The following section of this paper explains the model we built in order to obtain sufficient historical values for our prosumers, such that forming coalitions based on statistics is actually meaningfull.

@@

We now choose a very simple utility function that derives directly from the above remarks. If a coalition cannot provide a valid contract value, it receives naturally a utility of zero. Furthermore, it seems obvious that the utility should increase with the contract value. The $ 1/|S| $ term indicates that we favorise small coalitions, mainly because they are easier to maintain in terms of communications.

@@

The strategy of $ \{ij\} $ consists in maximizing its contract value as long as it respects this inequality, e.g to annonce $ P_{\phi}(ij) = \mu_{ij} - \sigma_{ij}\sqrt{2}erf^{-1}(1-2 \phi ) $. It thus appears (as it was intuitively understandable) that, for equivalent sizes, coalitions with low relative standard deviations ( $ \sigma_{ij} / \mu_{ij} $ ) are able to announce higher contract values.

@@

The model we used in order to simulate timeseries of available production provide some diversity because of the combo between the energetic mix and the climate vectors. Nevertheless, as the number of agents grows, the timeseries tend to exhibit similar patterns. This is particularly visible when creating a correlation graph $ G_{1}(\mathcal{A},E_{1},\omega_{1}) $ with the same kind of metric as \cite{Garas2008} or \cite{Onnela2004} ($ d_{ij} = 1 - \rho_{ij}^{2} $). Indeed, very well defined clusters appear in the corresponding $ \epsilon $-graph for any values of $ \epsilon $.

@@

We thus opt for reversing the metric ($ d_{ij}^{'} = \rho_{ij}^{2} $) such that uncorrelated timeseries are close to each other and (anti)correlated timeseries are distant in $ G_{2}(\mathcal{A},E_{2},\omega_{2}) $. As expected \cite{Onnela2004}, independently of the $ \epsilon $ parameter, the resulting $ \epsilon $-graph exhibits henceforth much less clustering than in $ G_{1} $ and communities seem hardly visible. Therefore, using classical clustering or community detection algorithms seem to provide poor results.

@@

In the following, we fix the reliability to a given empirical value ($\phi = Cste $) and observe how the coalitions evolves for different values of $P_{\phi}^{MIN}$. Note that the opposite is also possible although a little less intuitive, but not shown here because of space limitations and because it leads to similar conclusions.

On the right plot of figure \ref{Fig2}, we can see that when the grid constraints are fixed ($ \phi = 0.08 $ and $ P_{\phi}^{MIN} = 2 $ for instance) and only the number of coalitions varies, there is an increase in social welfare (sum of coalitions utilities) up to a maximum point (for $ N_{COAL} = 3 $ here) before a decrease. The reason is that increasing the number of coalitions allows more coalitions to reach stability and enter the market, but there is a point where nodes bringing stability are not sufficient inside the coalitions to make them pass the grid requirements, and some coalitions start to fail with zero utility. Moreover, reckon that increasing $ N_{COAL} $ means also increasing $ \epsilon $, leading to denser graphs where information is flooded, meaning that the algorithm performances also decrease.
Naturally, it is also visible that small values of $P_{\phi}^{MIN}$ lead to higher utilities because coalitions are able to announce higher contract values.

\begin{center}
\begin{figure}[b]
  \includegraphics[scale=0.8]{./figure5/coals.pdf}
  \caption{Evolution of the coalitions for different values of grid requirement $ P_{\phi}^{MIN} $. Blue dots represents clique percolation results while red dots stands for random process results.}
  \label{Fig3}
\end{figure}
\end{center}

For our 200 prosumers, we suppose that the grid fixes a reliability $ \phi = 0.1 $ (meaning that coalitions should produce more than their contract values at least $ 90 \% $ of the time). We also fix the number of coalitions to 10 and have a look at how these coalitions evolve as the grid changes the $ P_{\phi}^{MIN} $ requirement. As a comparison, we use a completely random algorithm that only asks for a number of coalitions and partitions the agents in a random fashion. Figure \ref{Fig3} shows this evolution for our algorithm (blue dots) and for the random process (red dots). The diameter of a dot is proportional to the number of agents in the coalition and the higher the dot is, the higher its utility. The $ P_{\phi}^{MIN} $ values of the x axis are expressed in tenth of MW for readability and the hatched zone corresponds to the "under-requirement space", meaning that whenever a coalition is in this zone, it has a null utility.

Looking at figure \ref{Fig3}, we can see first that our algorithm seems to make generally a better job at finding high valued coalitions, but also that the results seem far more robust against how the grid positions its requirements. The blue dots allowed to enter the market seem indeed to outnumber the red dots, especially when the grid requirements are neither to low or to high ($ P_{\phi}^{MIN} = 5 $ in figure \ref{Fig3}).

In more details, the left plot of figure \ref{Fig4} presents the evolution of social welfare as the number of coalitions increases (all other parameters are kept constant) for both random (red curve) and clique percolation (blue curve). As for the right plot, it shows the percentage of coalitions able to enter the market for different values of $ P_{\phi}^{MIN} $. For consistency, we average the results of both plots of the random procedure over 100 realizations (5 realizations for clique percolation, but the results are relatively stable), and the errorbars in the plots stands for the standard deviations of the results.

When the grid requirements are constant (left plot), and the number of desired coalitions is low, clique percolation generally performs only a little better than random. Nevetheless, when $ N_{COAL} $ gets bigger, the performance of a random split tumble down rapidely while clique percolation social welfare decreases slowly.

When the grid requirements vary, for very low $ P_{\phi}^{MIN} $ (close to zero), all coalitions for both algorithms are able to enter the market, yielding an acceptance percentage of $ 100 \% $. But as $ P_{\phi}^{MIN} $ increases, we see the percentage of the random procedure quickly dropping while it stays constant for our algorithm. For $ P_{\phi}^{MIN} = 4 $, we see that only a little more than half of the coalitions for the random case are able to enter while all of them enters for the clique percolation algorithm. Finally, when the grid requirement becomes to high, the acceptance percentage of our algorithm tends slowly to zero (only partly shown in this plot for readability).

\begin{center}
\begin{figure}
\setlength{\tabcolsep}{0em}
\begin{tabular}{cc}
   \includegraphics[scale=0.45]{./figure7/uti.eps} &
   \includegraphics[scale=0.45]{./figure4/survival.eps} \\
\end{tabular}
   \caption{Percentage of coalitions able to enter the market for different values of grid requirement $ P_{MIN} $. Blue curve represents clique percolation results while red curve stands for random process results}
   \label{Fig4}
\end{figure}
\end{center}
